---
title: "Study on impact of weather events in population health and economy of USA"
author: "Juan Sebastián Beleño Díaz"
date: "18 de noviembre de 2015"
output: html_document
---


### Synopsis

This is a study about what type of weather events produce more harmful consecuences in 
economy and population health in USA and tornados definitely are the most harmlful
weather event due to its big impact in injuries, fatalitites and property damage, 
their influence is concentrated in dates close to April and some of November. Some
weather events affects proportionally population health and economy, like floods,
tornados, TSTM winds, among others, but some event types affect more to population
health, like heat or excesive heat and this impact it's done during June (Summer),
others event types affect more to property damage like thunderstorm winds. The most
of the impact of weather events it's taken between March and October.

### Data Processing

To reach the results shown in the *synopsis* was necessary to process U.S. 
National Oceanic and Atmospheric Administration's (NOAA) storm database, 
available in <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>,
this file was already downloaded in this repository and it's possible to find it as 
`repdata-data-StormData.csv.bz2`.

It's mandatory to read and understand this dataset, for that reason It's loaded 
the data and it's printed to know how is structured this data. 

```{r readingdata, cache=TRUE}
stormData <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
str(stormData)
```

Among all this data I picked fatalities and injuries as key fields to get results 
about population health, and also I picked property damage as an indicator to show 
results about economy impact in weather events.

Taking this in mind, per each type of event  I processed the total of fatalities 
and injuries,and the total of property damage, with this data is possible to order 
the fields picked to represent population health and economy after a weather 
event happened:

```{r fieldsinorder, cache=TRUE, results='hide'}
# Injuries and Fatalities is to proccess health population consecuences in 
# weather events
injuriesPerTypeOfEvent <- aggregate(INJURIES ~ EVTYPE, data = stormData, sum)
fatalitiesPerTypeOfEvent <- aggregate(FATALITIES ~ EVTYPE, data = stormData, sum)

# Property damage is to proccess economy consecuences in weather events
propDamagePerTypeOfEvent <- aggregate(PROPDMG ~ EVTYPE, data = stormData, sum)

# Fields picked are ordered by event type
injuriesPerTypeOfEventOrdered <-
    injuriesPerTypeOfEvent[order(injuriesPerTypeOfEvent$INJURIES, 
                                 decreasing = TRUE),]
fatalitiesPerTypeOfEventOrdered<- 
    fatalitiesPerTypeOfEvent[order(fatalitiesPerTypeOfEvent$FATALITIES, 
                                   decreasing =TRUE),]

propDamagePerTypeOfEventOrdered <- 
    propDamagePerTypeOfEvent[order(propDamagePerTypeOfEvent$PROPDMG,
                                   decreasing = TRUE),]

# It's picked the top 5 of each field
TopFiveEventTypesByInjuries <- head(injuriesPerTypeOfEventOrdered, 5)
TopFiveEventTypesByFatalities <- head(fatalitiesPerTypeOfEventOrdered, 5)

TopFiveEventTypesByPropDMG <- head(propDamagePerTypeOfEventOrdered, 5)

```

Also to have a big picture about the behaivor of weather events impact during the time 
it's necesary to find the top 5 event type behaivor for each field picked and also it's 
a good information to know the behaivor per dates. This is done with this code:

```{r dataacrossthetime, cache=TRUE}
# I get the event types with more impact on health population and economy in U.S.
eventTypesWithMoreImpact <- c(
                                    as.character(TopFiveEventTypesByInjuries$EVTYPE),
                                    as.character(TopFiveEventTypesByFatalities$EVTYPE),
                                    as.character(TopFiveEventTypesByPropDMG$EVTYPE)
                            )
eventTypesWithMoreImpact <- unique(eventTypesWithMoreImpact)

# I subset the raw data just to contain the event types with most impact
stormDataWithMoreImpact <- stormData[stormData$EVTYPE %in% eventTypesWithMoreImpact,]

# It's created new fields to get years and month
stormDataWithMoreImpact$Datetime <- strptime( 
    stormDataWithMoreImpact$BGN_DATE, format='%m/%d/%Y  %H:%M:%S')
    stormDataWithMoreImpact$Month <- 
        as.numeric(format(stormDataWithMoreImpact$Datetime,'%m'))
    stormDataWithMoreImpact$Year <- 
        as.numeric(format(stormDataWithMoreImpact$Datetime,'%Y'))

# It's created new datasets separated by month
# I use sum instead of mean in this case because, there is a lack of data
# but this lack affect in the same way each month

injuriesPerEventTypeAndMonth <- aggregate(INJURIES ~ EVTYPE + Month, 
                                          data = stormDataWithMoreImpact,
                                          sum)
fatalitiesPerTypeOfEventAndMonth <- aggregate(  FATALITIES ~ EVTYPE + Month, 
                                                data = stormDataWithMoreImpact,
                                                sum)
propDamagePerTypeOfEventAndMonth <- aggregate(  PROPDMG ~ EVTYPE + Month, 
                                                data = stormDataWithMoreImpact,
                                                sum)

# It's created new datasets separated by year
# This time I use sum instead of mean because I want to see the behaivor
# using this way and see the lack of data in past years.

injuriesPerEventTypeAndYear <- aggregate(INJURIES ~ EVTYPE + Year, 
                                          data = stormDataWithMoreImpact,
                                          sum)
fatalitiesPerTypeOfEventAndYear <- aggregate(  FATALITIES ~ EVTYPE + Year, 
                                                data = stormDataWithMoreImpact,
                                                sum)
propDamagePerTypeOfEventAndYear <- aggregate(  PROPDMG ~ EVTYPE + Year, 
                                                data = stormDataWithMoreImpact,
                                                sum)

```

### Results

The following figure shows the top 5 of event types that produce more injuries, 
fatalities and property damage in the U.S. population during all the time in 
the data registered by NOAA, in these figures EVTYPE represents the weather 
event type and I think the others variables are clear.

```{r figurestopfive, cache=TRUE, fig.height=10, fig.width=8, fig.cap="Top five event types with impact in injuries, fatalities and economy of USA"}
    
library(gridExtra)
library(ggplot2)

# It's created a figure for each field
figureEventTypesByInjuries <- ggplot(data=TopFiveEventTypesByInjuries, 
                                     aes(x=EVTYPE, y=INJURIES))
figureEventTypesByInjuries <- figureEventTypesByInjuries + 
                              geom_bar(stat = 'identity') +
                              ggtitle("Top 5 event types with more injuries in U.S.")

figureEventTypesByFatalities <- ggplot(data=TopFiveEventTypesByFatalities, 
                                     aes(x=EVTYPE, y=FATALITIES))
figureEventTypesByFatalities <- figureEventTypesByFatalities + 
                                geom_bar(stat = 'identity') +
                            ggtitle("Top 5 event types with more fatalities in U.S.")

figureEventTypesByPropDMG <- ggplot(data=TopFiveEventTypesByPropDMG, 
                                     aes(x=EVTYPE, y=PROPDMG))
figureEventTypesByPropDMG <- figureEventTypesByPropDMG + 
                             geom_bar(stat = 'identity') +
                        ggtitle("Top 5 event types with more property damage in U.S.")

# It's shown the graphics in a panel
figuresTopFiveEventTypes <- list( figureEventTypesByInjuries, 
                                  figureEventTypesByFatalities, 
                                  figureEventTypesByPropDMG)

do.call(grid.arrange,figuresTopFiveEventTypes)

```


In the figure above it's obvious a big influence by far of tornados among any other 
weather event type in population health(injuries and fatalities) and 
economy(property damage), also we can appreciate that injuries are at least 5 
times less frequent that fatalities, some event types affects population health 
and economy proportionally as tornados, floods, TSTM winds among others, but 
also there are some events with implications in one of this indicators as heat 
and excesive heat that affect mostly the population health, but without cause 
property damage. 

Apart of figures I attached some tables with the top 10 of event type that causes 
more impact in economy and population health in U.S.

Here is the top 10 event types that produce more injuries in U.S.:

```{r tablestopteninjuries,results='asis', cache=TRUE}
library(xtable)
tableTopTenInjuriesPerEventType <- xtable(head(injuriesPerTypeOfEventOrdered, 10))
names(tableTopTenInjuriesPerEventType) <- c("Event Type", "Number of Injuries")
print(tableTopTenInjuriesPerEventType, 
      type="html",  
      html.table.attributes = list("border='1' align='center' width='500px' "))
```



Here is the top 10 event types that produce more fatalities in U.S.:

```{r tablestoptenfatalities,results='asis', cache=TRUE}
library(xtable)
tableTopTenFatalitiesPerEventType <- xtable(head(fatalitiesPerTypeOfEventOrdered, 10))
names(tableTopTenFatalitiesPerEventType) <- c("Event Type", "Number of Fatalities")
print(tableTopTenFatalitiesPerEventType, 
      type="html",  
      html.table.attributes = list("border='1' align='center' width='500px' "))
```



And finally here is the top 10 event types that produce more property damage in U.S.:

```{r tablestoptenpropdmg,results='asis', cache=TRUE}
library(xtable)
tableTopTenPropDMGPerEventType <- xtable(head(propDamagePerTypeOfEventOrdered, 10))
names(tableTopTenPropDMGPerEventType) <- c("Event Type", "Property damage")
print(tableTopTenPropDMGPerEventType, 
      type="html",  
      html.table.attributes = list("border='1' align='center' width='500px' "))
```

Also is possible to realize that some weather events happen during some periods of time,
so I think is a nice thing to see this data per year and see how the impact of weather 
events is changing during the time. 

```{r figuresperyear, cache=TRUE, fig.height=10, fig.width=8,fig.cap="Impact of weather events during the time"}
injuriesPerYear <- qplot(Year, 
                             INJURIES, 
                             data = injuriesPerEventTypeAndYear, 
                             geom = c("point", "line"), 
                             col = EVTYPE)
fatalitiesPerYear <- qplot(Year, 
                             FATALITIES, 
                             data = fatalitiesPerTypeOfEventAndYear, 
                             geom = c("point", "line"), 
                             col = EVTYPE)
propDamagePerYear <- qplot(Year, 
                             PROPDMG, 
                             data = propDamagePerTypeOfEventAndYear, 
                             geom = c("point", "line"), 
                             col = EVTYPE)

# Plot the data per year
figuresPerYear <- list(injuriesPerYear,
                           fatalitiesPerYear,
                           propDamagePerYear)

do.call(grid.arrange,figuresPerYear)
```

The very first thing we can see in this graphic is that exist some lack of data about 
some event types in years before 1995, the second thing is that in population health 
(injuries and fatalities) the event type with the most impact are tornados with some 
spikes of injuries because floods in 1998 and spikes in fatalities because heat and 
excesive heat in some recent years. The economy has a bunch of event types that 
affects it in apparently the same level of property damage, but in recents years, 
thunderstorm winds has dealt the major part of property damage.

Some weather events just happen in some parts of the year, so it's cool to see when
happen more weather events during a year. This could be seen in the next plot:

```{r figurespermonth, cache=TRUE, fig.height=10, fig.width=8, fig.cap="Impact of weather events in economy and population health distributed per months"}
injuriesPerMonth <- qplot(Month, 
                             INJURIES, 
                             data = injuriesPerEventTypeAndMonth, 
                             geom = c("point", "line"), 
                             col = EVTYPE)
fatalitiesPerMonth <- qplot(Month, 
                             FATALITIES, 
                             data = fatalitiesPerTypeOfEventAndMonth, 
                             geom = c("point", "line"), 
                             col = EVTYPE)
propDamagePerMonth <- qplot(Month, 
                             PROPDMG, 
                             data = propDamagePerTypeOfEventAndMonth, 
                             geom = c("point", "line"), 
                             col = EVTYPE)

# Plot the data per Month
figuresPerMonth <- list(injuriesPerMonth,
                           fatalitiesPerMonth,
                           propDamagePerMonth)

do.call(grid.arrange,figuresPerMonth)
```

I made this graphic because I know that U.S. has stations, so it's pretty important to 
point when weather events produce more impact in population health and economy.
Tornados has a big relevance in months close to April, but decrease after that 
with some minor spikes in November.

The others event types (excesive heat, flash floods, floods, heat, lightnings, 
thunderstorm winds and TSTM winds) have a raise close to June, with some 
exceptions like the increasing amount of injueries during October due to floods.