---
title: "Study on impact of weather events in population health and economy of USA"
author: "Juan Sebastián Beleño Díaz"
date: "18 de noviembre de 2015"
output: html_document
---


** Synopsis **

This is a study about what type of weather events produce more harmful consecuences in  economy and population health in USA.

** Data Processing **

To reach the results shown in the *synopsis* was necessary to process U.S. 
National Oceanic and Atmospheric Administration's (NOAA) storm database, 
available in <http://rmarkdown.rstudio.com>, this file was already downloaded 
in this repository and it's possible to find it as `repdata-data-StormData.csv`.

It's mandatory to read and understand this dataset, for that reason It's loaded 
the data and it's printed to know how is structured this data. 

```{r cache=TRUE}
stormData <- read.csv("repdata-data-StormData.csv")
str(stormData)
```

Among all this data I picked fatalities and injuries as key fields to get results 
about population health, and also I picked property damage as an indicator to show 
results about economy impact in weather events.

Taking this in mind, per each type of event  I calculate the number of fatalities 
and injuries,and the total of property damage, with this data is possible to order 
the fields picked to represent population health and economy after a weather event happened:

```{r cache=TRUE, results='hide'}
# Injuries and Fatalities is to proccess health population consecuences in weather events
injuriesPerTypeOfEvent <- aggregate(INJURIES ~ EVTYPE, data = stormData, sum)
fatalitiesPerTypeOfEvent <- aggregate(FATALITIES ~ EVTYPE, data = stormData, sum)

# Property damage is to proccess economy consecuences in weather events
propDamagePerTypeOfEvent <- aggregate(PROPDMG ~ EVTYPE, data = stormData, sum)

# Fields picked are ordered by event type
injuriesPerTypeOfEventOrdered <-
    injuriesPerTypeOfEvent[order(injuriesPerTypeOfEvent$INJURIES, 
                                 decreasing = TRUE),]
fatalitiesPerTypeOfEventOrdered<- 
    fatalitiesPerTypeOfEvent[order(fatalitiesPerTypeOfEvent$FATALITIES, 
                                   decreasing =TRUE),]

propDamagePerTypeOfEventOrdered <- 
    propDamagePerTypeOfEvent[order(propDamagePerTypeOfEvent$PROPDMG,
                                   decreasing = TRUE),]

# Gráfico de daños por meses
# Gráfico de daños por año
```

The following figure shows the top 5 of event types that produce more injuries, 
fatalities and property damage in the U.S. population during all the time in 
the data registered by NOAA.

```{r cache=TRUE}
# It's picked the top 5 of each field
TopFiveEventTypesByInjuries <- head(injuriesPerTypeOfEventOrdered, 5)
TopFiveEventTypesByFatalities <- head(fatalitiesPerTypeOfEventOrdered, 5)

TopFiveEventTypesByPropDMG <- head(propDamagePerTypeOfEventOrdered, 5)

# It's created a figure for each field
figureEventTypesByInjuries <- ggplot(data=TopFiveEventTypesByInjuries, 
                                     aes(x=EVTYPE, y=INJURIES))
figureEventTypesByInjuries <- figureEventTypesByInjuries + 
    geom_bar(stat = 'identity')

figureEventTypesByFatalities <- ggplot(data=TopFiveEventTypesByFatalities, 
                                     aes(x=EVTYPE, y=FATALITIES))
figureEventTypesByFatalities <- figureEventTypesByFatalities + 
    geom_bar(stat = 'identity')

figureEventTypesByPropDMG <- ggplot(data=TopFiveEventTypesByPropDMG, 
                                     aes(x=EVTYPE, y=PROPDMG))
figureEventTypesByPropDMG <- figureEventTypesByPropDMG + 
    geom_bar(stat = 'identity')

```
